---
title: "Exploring the `aids` Dataset"
author: Shuxian Fan
date: Sep 8, 2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description

The dataset used on this page is available in **R** as a data frame `aids` in the package **JM**. 
The dataset `aids` has 1408 observations measured with 9 variables shown as following:

| Variables  | Description |     
|----------|:----------------------------------------------|
| `patient` |  patients identiﬁer |
| `Time`|    time to death or censoring  | 
|`death` | a binary numerical variable with 0 denoting censoring and 1 death | 
|`CD4`| count of the CD4 cells|
|`obstime`| time at which CD4 cells count was recorded|
|`drug` | a factor with levels `ddC` denoting zalcitabine and `ddI` denoting didanosine|
|`gender`| a factor with levels `female` and `male`|
|`prevOI`| a factor with levels `AIDS` denoting previous AIDS diagnosis at study entry, and `noAIDS` denoting no previous infection.|
|`AZT`| a factor with levels `intolerance` and `failure` denoting AZT intolerance and AZT failure|

The corresponding dataset `aids.id` contains the first CD4 cell count for each patient. This dataset is used for survival analysis. 

### *Long* Format of the Data
The *long* format of the data requires the measurements of each subject are stored in multiple rows. It is in the contrast to the *wide* format where the measurements of each subject are stored in multiple columns. Here we would like to check if the dataset is already shaped in this format.

```{r load, message=FALSE}
# Load the package.
library(JM)
# Load the datasets
data("aids", package = "JM")
data("aids.id", package = "JM")
```

```{r pressure}
# Extract observations from patients id 1 to 3 with only variables: 
# "patient", "CD4", "obstime", "gender".
aids[aids$patient %in% c(1,2,3), c("patient", "CD4", "obstime", "gender")]
```
The measurements for the patients with id 1,2,3 are stored in the first 10 rows of the dataset, which agrees with the long format requirement.

```{r}
head(aids)
ncol(aids)
str(aids)
```


### Summary Statistics

```{r summary }
# Show the summary of the dataset `aid.id`
summary(aids.id)
```

The aids dataset has 467 unique patients, among which 45 are females, 422 are males. By the end of the study, there are in total 188 patients had died, giving a 40.26\% death rate.


## Data Exploration

### Longitudinal Submodel
Longitudinal data are usually fit using a mixed effects model. To determine the appropriate model used for the dataset, we would like to check the longitudinal trajectories for the longitudinal measurements. Here we pick a random group of 7 people, of which the CD4 cells counts are plot against the Time. 

```{r longitudinal_trajectories, message=FALSE}
# Randomly chose 8 patients' ids from the patient list.
set.seed(2)
index.spl = sample(aids.id$patient, 7)
# Extract the random sample from the dataset.
aids.spl1 = subset(aids, aids$patient %in% index.spl)

# Load ggplot2 package
require(ggplot2)
# Define base for the plot
b = ggplot(data = aids.spl1, aes(x = obstime, y = CD4, group = patient, color = patient))
# Plotting the trajectories
b+geom_line()+geom_point()


```

Based on the longitudinal trajectories, it is reasonable to consider a linear mixed effects model with random effects term in the intercept.
```{r, message=FALSE}
# install.packages(nlme)
library(nlme)
lmeFit = lme(CD4~obstime, random = ~1|patient, data = aids)
summary(lmeFit)
```

This model fit corresponds to a simple linear mixed effect model assuming a random effect term in the intercept for each patient:
$$y_{ij} = \beta_0 + \beta_{1}t_{ij}+b_{i}+\epsilon_{ij}$$
where $b_i$ and $\epsilon_{ij}$ are assumed to follow normal distributions.

Furthermore, we could compare the mean CD4 counts between genders. This will probably help us on answering the question that if the CD4 counts over time depend on whether a patient is female or male. Here we use 14 randomely chosen patients to illustrate the comparison. 

```{r}
# Get the 14 randomly selected sample.
set.seed(1)
males = aids.id[aids.id$gender=="male",]
females = aids.id[aids.id$gender!="male",]
male.id = sample(males$patient, 7)
female.id = sample(females$patient, 7)

# Extract the random sample from the dataset.
aids.spl2 = subset(aids, aids$patient %in% c(male.id, female.id))
# Define base for the plot
p = ggplot(data = aids.spl2, aes(x = obstime, y = CD4, group = patient, color = patient))
# Plot the Trajectories for each patients and adding the overall mean point.
p+geom_line()+geom_point() + stat_summary(aes(group = 1), geom = "point", fun.y = mean,
    shape = 16, size = 2.5)+facet_grid(.~gender)
```



### Survival Submodel
The survival data are characterized by the response variable being the time until the event(For example, death). Here for simplicity, we just illustrate the Kaplan-Meier estimate of the survival function for the `aids.id` dataset.
```{r surv}
# Create and plot the Kaplan-Meier estimate
km = survfit(Surv(Time, death)~1, data = aids.id)
plot(km, xlab = "Time", ylab = "Probability of Survival")
```

### Joint Modelling 
In order to answer the question:

**Whether there is any association between the longitudinal process and the hazard of death?**

We are interested in jointly fitting the longitudinal and survival submodels.

```{r, message=FALSE}
# install.packages("survival")
library(JM)
library(nlme)
library(survival)

# Joint model fit 
lmeFit = lme(CD4~obstime, random = ~1|patient, data = aids)
coxFit = coxph(Surv(Time,death)~1, data = aids.id, x = T)
jointFit = jointModel(lmeFit,coxFit, timeVar = "obstime", method = "piecewise-PH-aGH")
summary(jointFit)
```

The parameter of interest in the output of the joint model fitting is labeled `Assoct`. The output of the joint model indicates a strong association between the CD4 measurements and the risk of death. 

## Source and References

<http://www.biostat.umn.edu/~brad/data.html>

<http://www.biostat.umn.edu/~brad/software.html>

<https://stats.idre.ucla.edu/r/faq/how-can-i-visualize-longitudinal-data-in-ggplot2/>

